ProteInfer: deep networks for protein functional inference

### 2021-09-20 latex/draft3.tex
Various fixes

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for protein functional inference}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy Colwell}

@@

\begin{abstract}
Predicting the function of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. Here we instead employ deep convolutional neural networks to directly predict a variety of protein functions -- EC numbers and GO terms -- directly from an unaligned amino acid sequence. This approach provides precise predictions which complement alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}.
\end {abstract}

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We also examined the relationship between the number of examples of a label in the training dataset and the performance of the model. In an image recognition task, this is an important consideration since one image of, say, a dog, can be utterly different to another. Large numbers of labels are therefore required to learn filters that are able to predict members of a class. In contrast, for sequence data we found that even for labels that occurred less than five times in the training set, 58\% of examples in the test set were correctly recalled, while achieving a precision of 88\%, for an F1 of 0.7 (\Cref{sup:fig:ec_f1_by_training_size}%htmlver:<a href="#fig6" target="_self">Fig. 6</a>
). High levels of performance are maintained with few training examples because of the evolutionary relationship between sequences, which means that one ortholog of a gene may be similar in sequence to another. The simple BLAST implementation described above also performs well, and better than a single neural network, likely again exploiting the fact that many sequence have close neighbours in sequence space with similar functions. We again find that ensembling the BLAST and ProteInfer outputs provides performance exceeding that of either technique used alone.
%%%NUM_TRAIN_EGS%%%
\subsection*{Deep models link sequence regions to function}
%%%cam_colab%%% %
Proteins that use separate domains to carry out more than one enzymatic function are particularly useful in interpreting the behaviour of our model. For example, \textit{S. cerevisiae} fol1 (accession Q4LB35) catalyses three sequential steps of tetrahydrofolate synthesis, using three different protein domains (\Cref{fig:cam}A%htmlver:
). This protein is in our held-out test set, so no information about its labels was directly provided to the model.

@@

To visualise this space, we selected proteins with a single leaf-level EC number and used UMAP to compress their embeddings into two dimensions \citep{umap}.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{Figures/umap2.pdf}%
  \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
  \label{fig:umap}
\end{figure*}
%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

@@

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

@@

%startleaveout
\section*{Bibliography}
\bibliography{../bibliography}
\onecolumn
\newpage
%stopleaveout
\include{the_supplement}
%startleaveout

% Extra text

>>>>

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for protein functional inference}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell L. Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy J. Colwell}

@@

\begin{abstract}
  Predicting the function of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. Here we instead employ deep convolutional neural networks to directly predict a variety of protein functions -- EC numbers and GO terms -- directly from an unaligned amino acid sequence. This approach provides precise predictions which complement alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces, which we demonstrate with an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}
\end {abstract}

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}

\begin{figure*}[t]
  \centering
    \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
    % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
    \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
    \label{fig:cam}
  \end{figure*}

  \begin{figure*}
    \centering
      \includegraphics[width=\textwidth]{Figures/umap2.pdf}%
      \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
      \label{fig:umap}
    \end{figure*}

Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We also examined the relationship between the number of examples of a label in the training dataset and the performance of the model. In an image recognition task, this is an important consideration since one image of, say, a dog, can be utterly different to another. Large numbers of labels are therefore required to learn filters that are able to predict members of a class. In contrast, for sequence data we found that even for labels that occurred less than five times in the training set, 58\% of examples in the test set were correctly recalled, while achieving a precision of 88\%, for an F1 of 0.7 (\Cref{sup:fig:ec_f1_by_training_size}%htmlver:<a href="#fig6" target="_self">Fig. 6</a>
). High levels of performance are maintained with few training examples because of the evolutionary relationship between sequences, which means that one ortholog of a gene may be similar in sequence to another. The simple BLAST implementation described above also performs well, and better than a single neural network, likely again exploiting the fact that many sequence have close neighbours in sequence space with similar functions. We again find that ensembling the BLAST and ProteInfer outputs provides performance exceeding that of either technique used alone.
%%%NUM_TRAIN_EGS%%%

\subsection*{Deep models link sequence regions to function}
%%%cam_colab%%% %
Proteins that use separate domains to carry out more than one enzymatic function are particularly useful in interpreting the behaviour of our model. For example, \textit{S. cerevisiae} fol1 (accession Q4LB35) catalyses three sequential steps of tetrahydrofolate synthesis, using three different protein domains (\Cref{fig:cam}A%htmlver:
). This protein is in our held-out test set, so no information about its labels was directly provided to the model.

@@

To visualise this space, we selected proteins with a single leaf-level EC number and used UMAP to compress their embeddings into two dimensions \citep{umap}.

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

@@

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}

@@

%startleaveout
\begin{acknowledgements}
  We would like to thank Babak Alipanahi, Jamie Smith, Eli Bixby, Drew Bryant, Shanqing Cai, Cory McLean and Abhinay Ramparasad.
  \end{acknowledgements}
  \begin{contributions}
  TS was responsible for conception, data ingestion, code, analysis, visualisation, UI, paper writing. MLB was
  responsible for visualization, analysis, code, CLI, open sourcing, testing, paper writing. DB was responsible for
  paper writing, analysis of results, code reviews. LJC was responsible for problem conception and framing, analysis of results, paper writing.
  \end{contributions}

  \section*{Bibliography}
\bibliography{../bibliography}

\onecolumn
\newpage
%stopleaveout

%startleaveout
\section*{Supplement}
\beginsupplement

\subsection{Implementation details}

\subsubsection*{Label inheritance}
Some annotations, such as GO terms and EC numbers, are structured as directed acyclic graphics (which take the form of simple trees for EC numbers). Typically in such cases the annotation provided on UniProt is the most-specific that is known. For example if a protein is known to exhibit \textit{sequence specific DNA binding} (GO:0043565). It will not separately be annotated with the ancestral term \textit{DNA binding} (GO:0003677), this is simply assumed from the ontology. Using such an annotations directly, however, is likely to be problematic in a deep learning setting. Failing to annotate an example with the parental term demands that the model predict that the example is negative for this term, which is not the effect we want.
To address this our datasets include labels for all ancestors of applied labels for EC, GO and InterPro datasets. In the case of GO we restrict these to \textit{is\_a} relationships.

\subsubsection*{Class activation mapping}
Many proteins have multiple functional properties. For example we analyse the case of the bifunctional dhfr/ts of \textit{T. gondii}. Such bifunctional enzymes are often not unique -- it is functionally advantageous for these enzymes to be fused, which facilitates channeling of substrate between their active sites. Since there are a number of such examples within Swiss-Prot, the mere existence of a TS domain in a protein is (mild) evidence for possible DHFR function. To increase the interpretability of the network we subtract the activations for other predicted classes from those of the class of interest during class activation mapping.

\subsection*{Model architecture}
To create an architecture capable of receiving a wide range of input sequences, with computational requirements determined for each inference by the length of the individual input sequence, we employed a dilated convolutional approach \citep{dilatedconvolutions}. Computation for both training and prediction in such a model can be parallelized across the length of the sequence. By training on full length proteins, in a multi-label training setting, we aimed to build networks that could  extract functional information from raw amino acid sequences. One helpful feature of this architecture is its flexibility with regards to sequence length. Natural protein sequences can vary in length by at least three orders of magnitude, but some architectures have computational requirements that scale with the maximum sequence they are capable of receiving as input, rather than the sequence being currently examined. These fixed-length approaches reduce efficiency as well as place a hard limit on the length of sequences that can be examined.

\subsection{Input data statistics}

We use Swiss-Prot version 2019\_01 in our analysis, which gives us 559077 proteins, or 548264 after filtering for only 20 standard amino acids and filtering fragments.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
         Fold &  Number of sequences \\
\Xhline{2pt}
        train &               438522 \\
\hline
          dev &               55453 \\
\hline
         test &               54289 \\
\hline
 all together &               548264 \\
\hline
\end{tabular}
\caption{In our random split of the training data, we allocate about 80\% to the training fold, 10\% to the development fold, and 10\% to the test fold.}
\end{table}
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
         Fold &  Number of sequences \\
\Xhline{2pt}
        train &               182965 \\
\hline
          dev &               180309 \\
\hline
         test &               183475 \\
\hline
 all together &               546749 \\
\hline
\end{tabular}
\caption{In our clustered split of the training data, we use UniRef50, and allocate approximately equal numbers of sequences to each fold.}
\end{table}
Because different protein functions have differing prevalence, we note the number of proteins that have a given function for Pfam, EC, and GO labels, as well as noting the number of labels per protein.
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/go_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_random.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_random.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the random dataset.}
\end{figure}
\pagebreak
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the clustered dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the clustered dataset.}
\end{figure}
\pagebreak
When assigning examples to folds in our clustered dataset, we note that there are test examples that have labels that are never seen in the training data. We report these cases below as ``Impossible'' test example-label pairs.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &  Number \\
\Xhline{2pt}
                        Train labels &    3411 \\
\hline
                         Test labels &    3414 \\
\hline
              Impossible test labels &    1043 \\
\hline
           Train example-label pairs &  348105 \\
\hline
            Test example-label pairs &  348755 \\
\hline
 Impossible test example-label pairs &    3415 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for EC labels.}
\end{table}
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &   Number \\
\Xhline{2pt}
                        Train labels &    26538 \\
\hline
                         Test labels &    26666 \\
\hline
              Impossible test labels &     3739 \\
\hline
           Train example-label pairs &  8338584 \\
\hline
            Test example-label pairs &  8424299 \\
\hline
 Impossible test example-label pairs &    11137 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for GO labels.}
\end{table}
\pagebreak
\subsection{Precision/Recall curves}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{fig:with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{fig:with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{fig:with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{fig:with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{fig:with_ensemble} 
\end{figure}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_random_unzoomed.png}
    \caption{EC number prediction: random split} 
    \label{fig:without_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_random_unzoomed.png}
    \caption{GO term prediction: random split} 
    \label{fig:without_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_clustered_unzoomed.png}
    \caption{EC number prediction: clustered split} 
    \label{fig:without_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_clustered_unzoomed.png}
    \caption{GO term prediction: clustered split} 
    \label{fig:without_ensemble:d} 
  \end{subfigure} 
  \caption{Full precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs }
  \label{fig:without_ensemble} 
\end{figure}

\clearpage

\subsection{Stratified performance}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_f1_by_training_size.png}
  \caption{Performance of EC model stratified by number of training examples available for each test example.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}

\clearpage

\subsection{Hyperparameters}
We tuned over batch size, dilation rate, filters, first dilated layer, kernel size, learning rate, number of layers, mean vs max pooling, and Adam $\beta_1$, $\beta_2$ and $\epsilon$~\citep{kingma2014adam} over a number of studies to determine the set of parameters that optimized $\texttt{F}_{max}$. We found, as in~\cite{bileschi}, that the network was relatively unresponsive to slight changes in hyperparameters, and found that many of the hyperparameters that performed well in~\cite{bileschi} also performed well for this task. We chose to keep identical hyperparameters for the EC and GO tasks across both the random and clustered splits for simplicity, and we note that parameters with good performance on the random task performed respectively well on the clustered split. 
\begin{table}[htbp]
\centering
\def\arraystretch{1.1}
\begin{tabular}{|c|c|c|c|c|}
\hline
                           & \\
                           & CNN \\
\Xhline{1pt}
concurrent batches (data parallelism) & 8 \\
\hline
batch size                 & \shortstack{40 (per each GPU) \\ Dynamic based on sequence length} \\
\hline
dilation rate              & 3 \\
\hline
filters                    & 1100  \\
\hline
first dilated layer        & 2 \\
% zero-indexed so added one to the values shown in the hparams_registry for clarity.
\hline
gradient clip              & 1 \\
\hline
kernel size                & 9 \\
\hline
learning rate              & 1.5E-3 \\
\hline
learning rate decay rate   & 0.997 \\
\hline
learning rate decay steps  & 1000 \\
\hline
learning rate warmup steps & 3000 \\
\hline
Adam $\beta_1$              & .9 \\
\hline
Adam $\beta_2$              & .999 \\
\hline
Adam $\epsilon$             & 1E-8 \\
\hline
number of ResNet layers     & 5 \\
\hline
pooling                     & mean \\
\hline
ResNet bottleneck factor    & 0.5 \\
\hline
train steps                 & 500000 \\
\hline
\end{tabular}
\caption{Hyperparameters used in convolutional neural networks. We note that hyperparameters for single-GPU training are available in \url{github.com/google-research/proteinfer/blob/master/hparams_sets.py}.}
\label{tab:nn_hparams}
\end{table} 
\pagebreak
\subsection{Predicting coarse-grained functional localization with CAM}
The goal of this experimental methodology is to measure whether or not we correctly order the localization of function in bifunctional enzymes. As such, first we have to identify a set of candidates for experimentation.
